---
title: "PA1_template Assignment"
author: "Blessing Azubuike"
date: "2024-10-01"
output: html_document
---

## Loading required libraries

```{r load-libraries, message=FALSE, warning=FALSE}

library(dplyr)
library(ggplot2)
library(lubridate)
library(gridExtra)
```


## Loading and Preprocessing the Data

```{r load-data, echo=TRUE}
# Load the dataset
activity_data <- read.csv("C:\\Users\\bnnen\\Downloads\\repdata_data_activity\\activity.csv", stringsAsFactors = FALSE)

# Inspect the structure of the data
str(activity_data)

# Convert the date column to Date type
activity_data$date <- as.Date(activity_data$date, format="%Y-%m-%d")

```

## Mean Total Number of Steps Taken Per Day

```{r total-steps-per-day, echo=TRUE}
# Calculate the total number of steps taken per day, ignoring missing values
total_steps_per_day <- aggregate(steps ~ date, data = activity_data, sum, na.rm = TRUE)

# Plot a histogram of total steps taken per day
hist(total_steps_per_day$steps, breaks = 20, main = "Histogram of Total Steps per Day", 
     xlab = "Total Steps", col = "skyblue")

# Calculate mean and median of total steps per day
mean_steps <- mean(total_steps_per_day$steps)
median_steps <- median(total_steps_per_day$steps)

# Output the mean and median
mean_steps
median_steps

```


## Average Daily Activity Pattern

```{r daily-activity-pattern, echo=TRUE}
# Calculate the average number of steps for each 5-minute interval across all days
average_steps_interval <- aggregate(steps ~ interval, data = activity_data, mean, na.rm = TRUE)

# Time series plot of the average steps across intervals
plot(average_steps_interval$interval, average_steps_interval$steps, type = "l", 
     xlab = "5-minute Interval", ylab = "Average Steps", 
     main = "Average Daily Activity Pattern", col = "darkgreen")

```


## 5-Minute Interval with Maximum Steps

```{r max-steps-interval, echo=TRUE}
# Identify the interval with the maximum average steps
max_interval <- average_steps_interval[which.max(average_steps_interval$steps), ]
max_interval

```


## Imputing Missing Values

```{r impute-missing-values, echo=TRUE}
# Count the number of missing values in the dataset
total_na <- sum(is.na(activity_data$steps))

# Strategy: Fill missing values with the mean of the respective 5-minute interval
activity_data_filled <- activity_data

# Apply the average of the 5-minute interval to missing values only
mean_interval_steps <- ave(activity_data$steps, activity_data$interval, FUN = function(x) mean(x, na.rm = TRUE))
activity_data_filled$steps[is.na(activity_data_filled$steps)] <- mean_interval_steps[is.na(activity_data$steps)]

# Create a new dataset with imputed values
str(activity_data_filled)


```


## Histogram After Imputing Missing Values

```{r histogram-imputed, echo=TRUE}
# Calculate total steps per day for the dataset with imputed values
total_steps_imputed <- aggregate(steps ~ date, data = activity_data_filled, sum)

# Histogram of total steps taken per day after imputation
hist(total_steps_imputed$steps, breaks = 20, main = "Histogram of Total Steps per Day (Imputed)", 
     xlab = "Total Steps", col = "orange")

# Calculate mean and median of total steps per day after imputation
mean_steps_imputed <- mean(total_steps_imputed$steps)
median_steps_imputed <- median(total_steps_imputed$steps)

# Output the mean and median
mean_steps_imputed
median_steps_imputed

```

## Weekdays vs Weekends Activity Pattern

```{r weekdays-weekends, echo=TRUE}
# Create a new factor variable to classify weekdays and weekends
activity_data_filled$day_type <- ifelse(weekdays(activity_data_filled$date) %in% c("Saturday", "Sunday"), 
                                        "weekend", "weekday")

# Calculate the average steps per 5-minute interval across weekdays and weekends
average_steps_day_type <- aggregate(steps ~ interval + day_type, data = activity_data_filled, mean)

# Panel plot comparing weekday vs weekend activity patterns
library(ggplot2)
ggplot(average_steps_day_type, aes(x = interval, y = steps, color = day_type)) +
  geom_line() +
  facet_wrap(~ day_type, ncol = 1) +
  labs(title = "Average Steps per 5-Minute Interval: Weekdays vs Weekends", 
       x = "5-minute Interval", y = "Average Steps") +
  theme_minimal()

```


## Conclusion

This analysis provided insights into the activity patterns of the individual based on the dataset. The imputation of missing values slightly altered the summary statistics, but the overall patterns remained consistent. The differences between weekday and weekend activity patterns were observed, with distinct trends visible across the two categories.



